---
title: "deseq_analysis"
output: html_document
date: "2026-02-18"
---

Load libraries

```{r}
library(DESeq2)
library(tidyverse)
library(biomaRt)
library(apeglm)
```

Import count matrix

```{r}
counts <- read.csv( "counts_matrix.csv", row.names = 1)

colnames(counts) <- gsub("^X", "", colnames(counts))
colnames(counts) <- gsub("\\.", "-", colnames(counts))
colnames(counts) <- gsub("2O", "20", colnames(counts))

counts[is.na(counts)] <- 0
counts <- round(counts)

rownames(counts) <- gsub("\\..*", "", rownames(counts))
```

Create sample metadata

```{r}
sample_names <- colnames(counts)

coldata <- data.frame(
row.names = sample_names,
age = ifelse(grepl("young", sample_names), "young", "old")
)

coldata$age <- factor(coldata$age, levels = c("young","old"))

all(colnames(counts) == rownames(coldata))
```

DESeq2 analysis without protein-coding filtering

```{r}
dds_all <- DESeqDataSetFromMatrix(
countData = counts,
colData = coldata,
design = ~ age
)

dds_all <- dds_all[rowSums(counts(dds_all)) >= 20, ]

dds_all <- DESeq(dds_all)

res_all <- results(dds_all, contrast = c("age","old","young"))

res_all <- lfcShrink(
dds_all,
coef="age_old_vs_young",
type="apeglm"
)

res_df_all <- as.data.frame(res_all)
res_df_all$ensembl_gene_id <- rownames(res_df_all)
```

Add gene symbols

```{r}
mart <- useEnsembl(
biomart="genes",
dataset="mmusculus_gene_ensembl"
)

gene_map <- getBM(
attributes=c("ensembl_gene_id","mgi_symbol"),
filters="ensembl_gene_id",
values=res_df_all$ensembl_gene_id,
mart=mart
)

res_df_all <- merge(res_df_all,gene_map,by="ensembl_gene_id",all.x=TRUE)

res_df_all <- res_df_all %>%
mutate(
gene_label = ifelse(!is.na(mgi_symbol) & mgi_symbol!="",
mgi_symbol,
ensembl_gene_id)
)
```

Filter significant genes

```{r}
sig_all <- res_df_all %>%
filter(!is.na(padj) & padj < 0.05) %>%
select(gene_label,ensembl_gene_id,log2FoldChange,padj)
```

Expression filtering (>10 counts in ≥3 samples per group)

```{r}

counts_matrix <- counts(dds_all)

young_samples <- coldata$age == "young"
old_samples <- coldata$age == "old"

keep <- rowSums(counts_matrix[, young_samples] > 10) >= 3 &
        rowSums(counts_matrix[, old_samples] > 10) >= 3

dds_all <- dds_all[keep, ]

```

Genes upregulated in old

```{r}
sig_all_old <- res_df_all %>%
filter(!is.na(padj) & padj < 0.05 & log2FoldChange > 0) %>%
select(gene_label,ensembl_gene_id,log2FoldChange,padj) %>%
arrange(padj)
```

Genes upregulated in young

```{r}
sig_all_young <- res_df_all %>%
filter(!is.na(padj) & padj < 0.05 & log2FoldChange < 0) %>%
select(gene_label,ensembl_gene_id,log2FoldChange,padj) %>%
arrange(padj)
```

Export results for all genes

```{r}
write.csv(res_df_all,"DESeq2_results_all_genes.csv",row.names=FALSE)
write.csv(sig_all,"significant_genes_all.csv",row.names=FALSE)
write.csv(sig_all_old,"significant_genes_old_all.csv",row.names=FALSE)
write.csv(sig_all_young,"significant_genes_young_all.csv",row.names=FALSE)
```

Protein-coding gene filtering

```{r}
annotation <- getBM(
attributes=c("ensembl_gene_id","gene_biotype"),
filters="ensembl_gene_id",
values=rownames(counts),
mart=mart
)

protein_ids <- annotation %>%
filter(gene_biotype=="protein_coding") %>%
pull(ensembl_gene_id)

counts_pc <- counts[rownames(counts) %in% protein_ids, ]
```

DESeq2 analysis for protein-coding genes

```{r}
dds_pc <- DESeqDataSetFromMatrix(
countData=counts_pc,
colData=coldata,
design=~age
)

dds_pc <- dds_pc[rowSums(counts(dds_pc)) >= 20, ]

dds_pc <- DESeq(dds_pc)

res_pc <- results(dds_pc,contrast=c("age","old","young"))

res_pc <- lfcShrink(
dds_pc,
coef="age_old_vs_young",
type="apeglm"
)

res_df_pc <- as.data.frame(res_pc)
res_df_pc$ensembl_gene_id <- rownames(res_df_pc)

res_df_pc <- merge(res_df_pc,gene_map,by="ensembl_gene_id",all.x=TRUE)

res_df_pc <- res_df_pc %>%
mutate(
gene_label = ifelse(!is.na(mgi_symbol) & mgi_symbol!="",
mgi_symbol,
ensembl_gene_id)
)
```

Filter significant protein-coding genes

```{r}
sig_pc <- res_df_pc %>%
filter(!is.na(padj) & padj < 0.05) %>%
select(gene_label,ensembl_gene_id,log2FoldChange,padj)
```

Expression filtering for protein-coding genes

```{r}

counts_matrix <- counts(dds_pc)

young_samples <- coldata$age == "young"
old_samples <- coldata$age == "old"

keep <- rowSums(counts_matrix[, young_samples] > 10) >= 3 &
        rowSums(counts_matrix[, old_samples] > 10) >= 3

dds_pc <- dds_pc[keep, ]

```

Protein-coding genes upregulated in old

```{r}
sig_pc_old <- res_df_pc %>%
filter(!is.na(padj) & padj < 0.05 & log2FoldChange > 0) %>%
select(gene_label,ensembl_gene_id,log2FoldChange,padj) %>%
arrange(padj)
```

Protein-coding genes upregulated in young

```{r}
sig_pc_young <- res_df_pc %>%
filter(!is.na(padj) & padj < 0.05 & log2FoldChange < 0) %>%
select(gene_label,ensembl_gene_id,log2FoldChange,padj) %>%
arrange(padj)
```

Export protein-coding results

```{r}
write.csv(res_df_pc,"DESeq2_results_protein_coding.csv",row.names=FALSE)
write.csv(sig_pc,"significant_genes_protein_coding.csv",row.names=FALSE)
write.csv(sig_pc_old,"significant_genes_old_protein_coding.csv",row.names=FALSE)
write.csv(sig_pc_young,"significant_genes_young_protein_coding.csv",row.names=FALSE)
```